---
title: "ML_Project_DataCleaning"
author: "Siwei Wu"
date: "2023/3/3"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

set working directory and load packages
```{r}
# set up working directory
rm(list=ls())
path <- dirname(rstudioapi::getActiveDocumentContext()$path)
setwd(path)
rm(path)

# load packages
library(tidyverse)
library(dplyr)
library(RColorBrewer)
library(mice)
library(groupdata2)

```

## Functions for data cleaning
Function to Check NAs of each column in a dataframe
```{r}
# Check NAs
NA_count_fun = function(df){
  NA_count = sapply(df, function(x){sum(is.na(x))})
  NA_count = as.data.frame(NA_count)
  # NA_count = NA_count %>% filter(NA_count > 0)
  return(NA_count)
}

```

Function to Check NAs of each column of cases in a dataframe
```{r}
# Function to check NAs in cases
# df is the subset of the data with femalecancer_incident_case as the last column
NAInCases_count_fun = function(df){
  df = df %>% filter(femalecancer_incident_case==1)
  NA_in_cases = NA_count_fun(df)
  return(NA_in_cases)
}
```

Load in UKB data
```{r}
# read in data of exposure
ukb_cate = readRDS("./extraction_and_recoding/outputs/ukb_recoded_cate.rds")
ukb_numeric = readRDS("extraction_and_recoding/outputs/ukb_extracted_num.rds")
```

Deal with abnormal values in numeric variables
```{r}
# Check numeric variables with abnormal values(negative numbers for numeric variables)
abnormal_value = sapply(ukb_numeric, function(x){sum(ifelse(x<0, 1, 0), na.rm=TRUE)})
abnormal_value = as.data.frame(abnormal_value)
abnormal_value <- abnormal_value %>% filter (abnormal_value > 0)

# replace abnormal values with number or NA
  # -10 represents "Less than one"; -1 represents "Do not know"; -3 represents "Prefer not to answer"; -4 represents "Do not remember"
  # Substitute 0 to -10, NA to -1, -3 and -4
ukb_numeric[ukb_numeric==-10] = 0
ukb_numeric[ukb_numeric==-1] = NA
ukb_numeric[ukb_numeric==-3] = NA
ukb_numeric[ukb_numeric==-4] = NA
```

Deal with non-informative values in dataframe
```{r}
# set all "Do not know" and "Prefer not to answer" to NA
ukb_cate = ukb_cate %>% mutate_all(as.character)
ukb_cate[ukb_cate == "Do not know"] = NA
ukb_cate[ukb_cate == "Prefer not to answer"] = NA
```

Merge numeric data and categorical data
```{r}
ukb_cate["eid"] <- rownames(ukb_cate)
ukb_numeric["eid"] <- rownames(ukb_numeric)
ukb = inner_join(ukb_cate, ukb_numeric)

# Remove unuseful data
rm(abnormal_value)
rm(ukb_cate, ukb_numeric)
```

Merge the table of features with the table of outcome
```{r}
# Read in data of diabetes
T2DM <- readRDS("./outcome_definition/Outputs_T2DM/output_final.rds")
colnames(T2DM)<-c("eid", "T2DM_date_recr", "T2DM_diagnosis", "T2DM_date_death", "T2DM_case", "T2DM_prevalent_case", "T2DM_incident_case", "T2DM_time_to_diagnosis")

DM <- readRDS("./outcome_definition/Outputs_DM/output_final.rds")
colnames(DM)<-c("eid", "DM_date_recr", "DM_diagnosis", "DM_date_death", "DM_case", "DM_prevalent_case", "DM_incident_case", "DM_time_to_diagnosis")

HTN <- readRDS("./outcome_definition/Outputs_HTN/output_final.rds")
colnames(HTN)<-c("eid", "HTN_date_recr", "HTN_diagnosis", "HTN_date_death", "HTN_case", "HTN_prevalent_case", "HTN_incident_case", "HTN_time_to_diagnosis")

Dyslipid <- readRDS("./outcome_definition/Outputs_Dyslipid/output_final.rds")
colnames(Dyslipid)<-c("eid", "Dyslipid_date_recr", "Dyslipid_diagnosis", "Dyslipid_date_death", "Dyslipid_case", "Dyslipid_prevalent_case", "Dyslipid_incident_case", "Dyslipid_time_to_diagnosis")


ukb <- inner_join(ukb, (T2DM  %>% select(eid, T2DM_prevalent_case, T2DM_incident_case)))
ukb <- inner_join(ukb, (DM  %>% select(eid, DM_prevalent_case, DM_incident_case)))

ukb <- inner_join(ukb, (HTN  %>% select(eid, HTN_prevalent_case)))
ukb <- inner_join(ukb, (Dyslipid %>% select(eid, Dyslipid_prevalent_case)))

rm(T2DM, DM, HTN, Dyslipid)

```

Cohort construction
```{r}
# Remove prevalent cases (diagnosed before enrollment)
ukb <- ukb %>% filter(DM_prevalent_case == 0) # remove prevalent cases
ukb <- ukb %>% select(-DM_prevalent_case, -T2DM_prevalent_case)
table(ukb$DM_incident_case) # 32252 Incident DM diabetes
table(ukb$T2DM_incident_case) #30539 Incident T2DM diabetes

# Remove fasting glucose > 7, HbA1c >= 42mmol/mol
ukb = ukb %>% filter(Bio_glucose.0.0<7)
ukb = ukb %>% filter(Bio_glycated_haemoglobin.0.0<42)

# Remove non-white people
ukb <- ukb %>% filter(Demo_Race.0.0=="British"|Demo_Race.0.0=="Irish"|Demo_Race.0.0=="Any other white background"|Demo_Race.0.0=="White") # only keep white people
ukb = ukb %>% select(-Demo_Race.0.0)
```


## Data Cleaning
Clean family history illness data
```{r}
# Illness_mother
ukb = ukb[!is.na(ukb$illness_mother.0.0),] # Exclude NAs
ukb_illness_mother = ukb %>% select(starts_with("illness_mother"))
ukb["Fam_Mother_DM"] = apply(ukb_illness_mother, MARGIN = 1, FUN= function(x) all("Diabetes" %in% x))
ukb = ukb %>% select(-colnames(ukb_illness_mother))

# Illness_father
ukb = ukb[!is.na(ukb$illness_father.0.0),] # Exclude NAs
ukb_illness_father = ukb %>% select(starts_with("illness_father"))
ukb["Fam_Father_DM"] = apply(ukb_illness_father, MARGIN = 1, FUN= function(x) all("Diabetes" %in% x))
ukb = ukb %>% select(-colnames(ukb_illness_father))

# Illness_siblings.0.0: we won't use siblings' diabetes history, so don't clean this variable
# ukb = ukb[!is.na(ukb$illness_siblings.0.0),] # Exclude NAs
# ukb_illness_siblings = ukb %>% select(starts_with("illness_siblings"))
# ukb["Fam_siblings_DM"] = apply(ukb_illness_siblings, MARGIN = 1, FUN= function(x) all("Diabetes" %in% x))
# ukb = ukb %>% select(-colnames(ukb_illness_siblings))

# Remove redundant 
rm(ukb_illness_father, ukb_illness_mother)
```

Clean lifestyle data
```{r}
# Factor 
ukb$Life_Alcohol_freq.0.0 = factor(ukb$Life_Alcohol_freq.0.0, levels=c("Never", "Special occasions only", "Once or twice a week", "One to three times a month", "Three or four times a week", "Daily or almost daily"))
ukb$Life_Curr_Smoking.0.0 = factor(ukb$Life_Curr_Smoking.0.0 , levels=c( "No", "Only occasionally", "Yes, on most or all days"))

# Remove redundant smoking and alcohol variables
ukb = ukb %>% select(-c("Life_smok_status.0.0", "Life_Alcohol_status.0.0", "Life_ever_smoked.0.0"))
rm(ukb_life, NA_count_ukb_life)
```

Clean biochemical and urine variables
```{r}
# Biochemistry and urine variables: Clean the column of Urine_microalbumin
# Urine_microalbumin < 6.7 is flagged, set people with Urine_microalbumin < 6.7 to be 6.7 in their Urine_microalbumin
ukb = ukb %>% mutate(Urine_microalbumin.0.0 = ifelse(is.na(Urine_microalbumin_flag.0.0), Urine_microalbumin.0.0, 6.7)) %>% select(-Urine_microalbumin_flag.0.0)

# Remove Oestradiol and Rheumatoid
ukb = ukb %>% select(-c("Bio_Oestradiol.0.0", "Bio_Rheumatoid factor.0.0"))
```

Clean physical measurement variables
```{r}
# Check NAs in physical measurement variables 
ukb_PhyMeas = ukb %>% select(starts_with("phyMeas_"))
NA_count_ukb_PhyMeas = NA_count_fun(ukb_PhyMeas)
# table(ukb_PhyMeas$femalecancer_incident_case) # 12536 cases

# Clean blood pressure: Fill in NAs in auto-read blood pressure with the one manually measured
ukb = ukb %>% mutate(phyMeas_SBP = ifelse(is.na(PhyMeas_SBP_auto.0.0), PhyMeas_SBP_manual.0.0, PhyMeas_SBP_auto.0.0))
ukb = ukb %>% mutate(phyMeas_DBP = ifelse(is.na(PhyMeas_DBP_auto.0.0), PhyMeas_DBP_manual.0.0, PhyMeas_DBP_auto.0.0))
ukb = ukb %>% select(-c("PhyMeas_SBP_manual.0.0", "PhyMeas_SBP_manual.0.1", "PhyMeas_DBP_manual.0.0", "PhyMeas_DBP_manual.0.1", "PhyMeas_DBP_auto.0.0", "PhyMeas_DBP_auto.0.1", "PhyMeas_SBP_auto.0.0","PhyMeas_SBP_auto.0.1"))


# Drop observations with NAs in physical measurement variables
# ukb_PhyMeas = ukb %>% select(starts_with("phyMeas_"))
# ukb = ukb[complete.cases(ukb_PhyMeas),]
rm(ukb_PhyMeas, NA_count_ukb_PhyMeas, Unknown_count_ukb_PhyMeas)

```

Clean diet data
```{r}
# str(ukb_ML)
# lapply(ukb_ML %>% select(starts_with("diet")), table)

# Factor ordinal diet data
for(diet in c("Diet_oil_fish_intake.0.0", "Diet_non_oil_fish_intake.0.0", "Diet_proc_meat_intake.0.0", "Diet_poultry_intake.0.0", "Diet_beef_intake.0.0", "Diet_mutton_intake.0.0", "Diet_pork_intake.0.0")){ukb[,diet] = factor(ukb[,diet], levels=c("Never", "Less than once a week", "Once a week", "2-4 times a week", "5-6 times a week", "Once or more daily"))}

# Factor ordinal diet data: Diet_added_salt.0.0
ukb$Diet_added_salt.0.0  = factor(ukb$Diet_added_salt.0.0 , levels=c("Never/rarely","Sometimes", "Usually", "Always"))
```

Remove columns with too many NAs(more then 50%)
```{r}
ukb = ukb %>% select(-starts_with("Nutrient_"))
ukb = ukb %>% select(-starts_with("Med_Treat"))
ukb = ukb %>% select(-starts_with("Meta_"))
ukb = ukb %>% select(-starts_with("illness_siblings"))
ukb = ukb %>% select(-starts_with("Diet_"))
ukb = ukb %>% select(-T2DM_incident_case)
```

Keep complete cases
```{r}
ukb_ml = ukb[complete.cases(ukb),]
table(ukb_ml$DM_incident_case)
```

Factor variables
```{r}
# Select factor variables and 
ukb_ml_factor= ukb %>% select(eid, where(is.factor))

# Transform factors into ordinal variables
ukb_ml_ordinal = as.data.frame(lapply(ukb_ml_factor, as.numeric))
ukb_ml_ordinal$eid = as.character(ukb_ml_ordinal$eid)
  
# Select non factor variables
ukb_ml_nonfactor = ukb_ml %>% select(-is.factor)
# One hot code "Demo_Sex.0.0", "DM_incident_case", "HTN_prevalent_case", "Dyslipid_prevalent_case", "Fam_Mother_DM", "Fam_Father_DM"   
# ukb_ml_nonfactor$Fam_Mother_DM = as.factor(as.numeric(ukb_ml_nonfactor$Fam_Mother_DM))
# ukb_ml_nonfactor$Fam_Father_DM = as.factor(as.numeric(ukb_ml_nonfactor$Fam_Father_DM))
ukb_ml_nonfactor$Fam_Mother_DM = as.numeric(ukb_ml_nonfactor$Fam_Mother_DM)
ukb_ml_nonfactor$Fam_Father_DM = as.numeric(ukb_ml_nonfactor$Fam_Father_DM)
# ukb_ml_nonfactor$HTN_prevalent_case = as.factor(ukb_ml_nonfactor$HTN_prevalent_case)
# ukb_ml_nonfactor$Dyslipid_prevalent_case = as.factor(ukb_ml_nonfactor$Dyslipid_prevalent_case )
ukb_ml_nonfactor$Demo_Sex.0.0 = ifelse(ukb_ml_nonfactor$Demo_Sex.0.0 == "Female", 0, 1)
# ukb_ml_nonfactor$Demo_Sex.0.0 = factor(ukb_ml_nonfactor$Demo_Sex.0.0, levels=c(0,1))

# Merge
ukb_ml = inner_join(ukb_ml_nonfactor, ukb_ml_ordinal, by="eid")
rm(ukb_ml_factor, ukb_ml_ordinal, ukb_ml_nonfactor)
```

Select data for predicting
```{r}
# ukb_ML = ukb %>% select(eid, Demo_Sex.0.0, Demo_Age_enrollment.0.0, PhyMeas_waist_cir.0.0, PhyMeas_BMI.0.0, Fam_Par_DM, Bio_hdl_cholesterol.0.0, Bio_triglycerides.0.0, starts_with("Diet_"), HTN_prevalent_case, Dyslipid_prevalent_case, DM_incident_case)
# ukb_ML = ukb_ML[complete.cases(ukb_ML),]
# ukb_ML_NA_count = NA_count_fun(ukb_ML)
# table(ukb_ML$DM_incident_case)


## Rearrange the variables
sex_horm = c("Bio_shbg.0.0","Bio_testosterone.0.0")
Inflam = c("Bio_c_reac_protein.0.0")
glucose = c("Bio_glucose.0.0", "Bio_glycated_haemoglobin.0.0", "Bio_igf_1.0.0")
liver = c("Bio_alkaline_phosphate.0.0", "Bio_alanine_aminotransferase.0.0", "Bio_aspartate_aminotransferase.0.0", "Bio_direct_bilirubin.0.0", "Bio_gamma_glutamyltransferase.0.0", "Bio_total_bilirubin.0.0")
lipid = c("Bio_apolipoprotein_a.0.0", "Bio_apolipoprotein_b.0.0", "Bio_cholesterol.0.0", "Bio_hdl_cholesterol.0.0", "Bio_ldl_direct.0.0", "Bio_lipoprotein_a.0.0", "Bio_triglycerides.0.0")
nutrit = c("Bio_albumin.0.0", "Bio_total_protein.0.0")
skeleton = c("Bio_calcium.0.0", "Bio_phosphate.0.0", "Bio_vitamin_d.0.0")
kidney = c("Bio_urea.0.0", "Bio_creatinine.0.0", "Bio_cystatin_c.0.0")
PhyMeas = c("phyMeas_SBP", "phyMeas_DBP", "PhyMeas_waist_cir.0.0", "PhyMeas_BMI.0.0")
life = c( "Life_Days_WK_PhyAc.0.0", "Life_Curr_Smoking.0.0", "Life_Alcohol_freq.0.0")

ukb_ml = ukb_ml %>% select(eid, starts_with("Demo_"), PhyMeas, life, sex_horm, Inflam, glucose, liver, lipid, nutrit, skeleton, Bio_urate.0.0, kidney, starts_with("Urine_"), HTN_prevalent_case, Dyslipid_prevalent_case, Fam_Mother_DM, Fam_Father_DM, DM_incident_case)

rm(PhyMeas, life, sex_horm, Inflam, glucose, liver, lipid, nutrit, skeleton, kidney)
table(ukb_ml$DM_incident_case)
```


```{r}
colnames(ukb_ml) = c("eid", "Sex", "Age", "SBP", "DBP", "Waist_Cir", "BMI", "Phy_Activity", "Smoking", "Alcohol", 
  "shbg", "Testosterone", "CRP", "Glucose", "HbA1c", "IGF1", "ALP", "ALT", "AST", 
  "Direct_bilirubin", "GGT", "Total_bilirubin", "ApoA", "ApoB", 
  "Cholesterol", "HDLC", "LDLC", "Lipoprotein_A", "Triglycerides", 
  "Albumin", "Total_protein", "Calcium", "Phosphate", "Vitamin_D", 
  "urate", "Urea", "Creatinine", "Cystatin_C", 
  "Urine_Microalbumin", "Urine_Creatine", "Urine_Potassium", "Urine_Sodium",
  "HTN", "Dyslipid", "DM_Mother", "DM_Father", "DM")
```



## Data preparation
Split test set and training set 
```{r}
set.seed(123)
rownames(ukb_ml)=ukb_ml$eid
ukb_ml = ukb_ml %>% select(-eid)
train_test_split=createDataPartition(ukb_ml$DM, p = .75, list = FALSE, times = 1)
ukb_ml_train=ukb_ml[train_test_split, ]
ukb_ml_test=ukb_ml[-train_test_split, ]
rm(train_test_split)
table(ukb_ml_train$DM_incident_case)
table(ukb_ml_test$DM_incident_case)
```

Down-sample control to a case:control = 1:5
```{r}
# plot(density(ukb_ml$Age))
# lines(ukb_ml$Age)
# Most of our obsevations are between 40 - 70 years old
# The prevalence of diabetes between 45 - 75 ranges from 10% - 17%
# https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/612306/Diabetesprevalencemodelbriefing.pdf

# keep case: control to be 1:5, so that the ration of case and control assimilate the real world one

ratio = 5 # ratio = case: control
## down-sample ukb_ml_train
ids_case = which(ukb_ml_train$DM == 1)
ids_control = which(ukb_ml_train$DM == 0)
set.seed(100)
ids_control_downsample = sample(ids_control, ratio*length(ids_case))
ukb_ml_train = ukb_ml_train[sort(c(ids_case, ids_control_downsample)),]
table(ukb_ml_train$DM)

## down-sample ukb_ml_test 
# ids_case = which(ukb_ml_test$DM_incident_case == 1)
# ids_control = which(ukb_ml_test$DM_incident_case == 0)
# set.seed(100)
# ids_control_downsample = sample(ids_control, ratio*length(ids_case))
# ukb_ml_test = ukb_ml_test[sort(c(ids_case, ids_control_downsample)),]
# table(ukb_ml_test$DM_incident_case)
```

Scaled version
```{r}
X_train = ukb_ml_train %>% select(-DM) 
X_train = as.matrix(X_train)
X_train = scale(X_train, center=TRUE, scale=TRUE)
print(all(rownames(X_train)==rownames(ukb_ml_train)))
ukb_ml_train_scale = cbind(X_train, ukb_ml_train %>% select(DM))
rm(X_train)


X_test = ukb_ml_test %>% select(-DM)
X_test = as.matrix(X_test)
X_test = scale(X_test, center=TRUE, scale=TRUE)
print(all(rownames(X_test)==rownames(ukb_ml_test)))
ukb_ml_test_scale = cbind(X_test, ukb_ml_test %>% select(DM))

write.csv(ukb_ml_train_scale, "ML_train_scale.csv", row.names=FALSE)
write.csv(ukb_ml_test_scale, "ML_test_scale.csv", row.names=FALSE)
```
Un-scaled version(Binary variables are factors, but csv will not memorize the data type of factor; therefore need to re-factor when analyzing)
```{r}
# ukb_ml_train$Demo_Sex.0.0 = factor(ukb_ml_train$Demo_Sex.0.0)
# ukb_ml_train$HTN_prevalent_case = factor(ukb_ml_train$HTN_prevalent_case)
# ukb_ml_train$Dyslipid_prevalent_case = factor(ukb_ml_train$Dyslipid_prevalent_case)
# ukb_ml_train$Fam_Mother_DM = factor(ukb_ml_train$Fam_Mother_DM)
# ukb_ml_train$Fam_Father_DM = factor(ukb_ml_train$Fam_Father_DM)
# ukb_ml_train$DM_incident_case = factor(ukb_ml_train$DM_incident_case)
# 
# ukb_ml_test$Demo_Sex.0.0 = factor(ukb_ml_test$Demo_Sex.0.0)
# ukb_ml_test$HTN_prevalent_case = factor(ukb_ml_test$HTN_prevalent_case)
# ukb_ml_test$Dyslipid_prevalent_case = factor(ukb_ml_test$Dyslipid_prevalent_case)
# ukb_ml_test$Fam_Mother_DM = factor(ukb_ml_test$Fam_Mother_DM)
# ukb_ml_test$Fam_Father_DM = factor(ukb_ml_test$Fam_Father_DM)
# ukb_ml_test$DM_incident_case = factor(ukb_ml_test$DM_incident_case)

write.csv(ukb_ml_train, "ML_train.csv", row.names=FALSE)
write.csv(ukb_ml_test, "ML_test.csv", row.names=FALSE)
```

